---
title: "Basic answers about severe weather events, using the NOAA Storm Database"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

## 1. Synopsis

*(This page is my work for the *Peer-graded Assignment (Course Project 2)* of the "Reproducible Research" course by Johns Hopkins University, tought on Coursera.)*

The basic goal of this report is to answer some basic questions about severe weather events. We use the NOAA (National Oceanic and Atmospheric Administration) database to answer the questions below, showing the code for the entire analysis.

The questions are:

- Across the United States, which types of events are most harmful with respect to population health?
- Across the United States, which types of events have the greatest economic consequences?

## 2. Environment Setup

Here we import the needed lib(s), and set the chunk options globally for this page:
```{r init, echo=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE)
```

The following information may be useful to consistantly reproduce this study:
```{r session-info, echo=TRUE}
sessionInfo()
```

## 3. Data Processing
### 3.1 Loading and preprocessing the data

Let's unpack the compressed archive and load the enclosed CSV data. We will not bother parsing dates.
```{r data-load, cache=TRUE}
file_name <- "repdata_data_StormData.csv.bz2"
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if (!file.exists(file_name)) {
    download.file(url, file_name)
}

stormdata <- read.csv(file_name) # decompress and load
```

### 3.2 Examining the event types
As we can see in Annex A, the EVTYPE variable is not normalized, and there are also typos.

### 3.3 Grouping event types
In order to have more synthetic event categories (less groups and no typo effects) let's create the `evgroup` variable:
```{r data-group, cache=TRUE}
evgroup <- tolower(as.character(stormdata$EVTYPE)) # copy original events as lowercase

evgroup[grepl("\\<(blizzard)\\>", evgroup)] <- "BLIZZARD"
evgroup[grepl("\\<(cold|cool|low temp(erature)?|hypothermia|windchill)\\>", evgroup)] <- "COLD"
evgroup[grepl("\\<(dry(ness)?|driest|drought)\\>", evgroup)] <- "DROUGHT"
evgroup[grepl("\\<(erosio?n|landslump)\\>", evgroup)] <- "EROSION"
evgroup[grepl("\\<(wild)?fires?\\>", evgroup)] <- "FIRE"
evgroup[grepl("\\<(flo+d(|s|ing)|fldg|stream|drowning|(coastal)?flood(|in))\\>", evgroup)] <- "FLOOD"
evgroup[grepl("\\<(fog|dust(storm)?|smoke|)\\>", evgroup)] <- "FOG_AND_DUST"
evgroup[grepl("\\<hail(storms?)?\\>", evgroup)] <- "HAIL"
evgroup[grepl("\\<(warm|warmth|hot|high temperatures?|heat(burst)?|hyperthermia)\\>", evgroup)] <- "HEAT"
evgroup[grepl("(hurricane|typhoon|tropical depression)", evgroup)] <- "HURRICANE"
evgroup[grepl("\\<ic[ye]|freeze|freezing|frost|glaze|sleet\\>", evgroup)] <- "ICE"
evgroup[grepl("\\<lig[hn]tn?ing\\>", evgroup)] <- "LIGHTNING"
evgroup[grepl("\\<(precip([ia]tation)?|rainfall|rains?|showers?|rainstorms?)\\>", evgroup)] <- "PRECIPITATION"
evgroup[grepl("slide", evgroup)] <- "SLIDE"
evgroup[grepl("\\<((thunder)?snow(storm)?|snow(s|fall|pack)?|avalanch?e)\\>", evgroup)] <- "SNOW"
evgroup[grepl("\\<((coastal)?storm|thun?d?e+r?s?torm[sw]?|thunderestorm|thunderstrom|tstmw?)", evgroup)] <- "THUNDERSTORM"
evgroup[grepl("\\<((tornado|TORNDAO)e?s?|funnels?|gustnado|microburst|landspout|wall cloud|downburst|way?ter(spout.*)?)\\>", evgroup)] <- "TORNADO"
evgroup[grepl("\\<(volcanic|vog)\\>", evgroup)] <- "VOLCANISM"
evgroup[grepl("\\<(tsunami|surf|currents?|rogue wave|seas?|marine|tides?|high waves|swells|coastal surge)\\>", evgroup)] <- "WATER"
evgroup[grepl("\\<wet(ness)?\\>", evgroup)] <- "WETNESS"
evgroup[grepl("\\<(wind|wnd|(whirl)?winds?)\\>", evgroup)] <- "WIND"
evgroup[grepl("\\<(winter|winte?ry)\\>", evgroup)] <- "WINTER"

evgroup[grepl("[a-z]", evgroup)] <- "OTHER" # collect any uncategorized original events (lowercase)

evgroup <-as.factor(evgroup)
```
This was quite some work, and we'll see later if it pays.

All events that were not categorized are collected all together in the `OTHER` group. We will see if doing so has only marginal effect on the results.

### 3.4 Exploring the data
Let's evaluate the impact on health (fatalities and injuries):
```{r data-explore, cache=TRUE}
health.impact <- stormdata$FATALITIES + stormdata$INJURIES;
df = data.frame(evgroup=evgroup, value=health.impact)
impact <- aggregate(value ~ evgroup, data = df, sum)
impact.ordered.top <- impact[order(impact$value, decreasing = TRUE),][1:10,]
```

Impact per natural event group:
```{r data-explore-2, cache=TRUE}
impact.ordered.top
```
As we can see, there is a domination of the first event over the others, by almost one order of magnitude. So to better distinguish the values, we will plot the values with a logarithmic scale. 

### 3.5 Plotting helper
At this point, as we plan to produce a few aggregated graphs and lists of identical nature, let's create a function to generate them easily later:
```{r helper-plot, cache=TRUE}
makeplot <- function(events, values, what, ylab) {
    top.n <- 10;
    df = data.frame(evgroup=events, value=values)
    impact <- aggregate(value ~ evgroup, data = df, sum)
    impact.ordered.top <- impact[order(impact$value, decreasing = TRUE),][1:top.n,]
    par(mar = c(11,4,4,4), cex = 0.75)
    barplot(impact.ordered.top$value,
        main = paste("Impact of natural events on", what),
        ylab = paste(ylab, "(log)"),
#        xlab = paste(ylab, "caused by natural events"),
        names.arg = impact.ordered.top$evgroup,
        cex.axis = 0.75,
        log = "y",  # log scale!
        las = 2)
}

make_list <- function(events, values) {
    top.n <- 10;
    df = data.frame(evgroup=events, value=values)
    impact <- aggregate(value ~ evgroup, data = df, sum)
    impact.ordered.top <- impact[order(impact$value, decreasing = TRUE),][1:top.n,]
    print(impact.ordered.top)
}
```

To quantify the damages, two variables (per damage) are used: the value, and an exponent in numeric or alphabetic form.
Let's create a function to compute the numeric damage value:
```{r helper-decode-value, cache=FALSE}
decode_value <- function(val, exp) {
    res <- val
    e <- tolower(exp)
    res[e=="b"] <-res[e=="b"] * 1e9
    res[e=="m"] <-res[e=="m"] * 1e6
    res[e=="k"] <-res[e=="k"] * 1e3
    res[e=="2"] <-res[e=="2"] * 1e2
    res
}
```

## 4. Results
### 4.1 Impact of natural events on population health
```{r impact-health, cache=FALSE}
health.impact <- stormdata$FATALITIES + stormdata$INJURIES;
makeplot(evgroup, health.impact, "health", "Fatalities and injuries")
```

We see a clear dominant impact of tornados.

### 4.2 Impact on property damage
```{r impact-property, cache=FALSE}
property.damages <- decode_value(stormdata$PROPDMG, stormdata$PROPDMGEXP)
makeplot(evgroup, property.damages, "property damage", "Financial damage")
```

Floods dominate, followed by hurricanes, thunderstorms and tornados.

### 4.3 Impact on crop damage
```{r impact-crop, cache=FALSE}
crop.damages <- decode_value(stormdata$CROPDMG, stormdata$CROPDMGEXP)
makeplot(evgroup, crop.damages, "crop damage", "Ĉrop damage")
```

Droughts and floods dominate.

## 5. Further Research
To strictly restrict the research to the United States, geographic data enclosed in the data set could be used to filter the observations.


HURRICANE clearly dominate.

Comment: crops looks more resistant to natural events (other than HURRICANE) than property items.

# Annex A - Event types
The EVTYPE variable of the dataset is:
```{r data-examine, cache=TRUE}
levels(stormdata$EVTYPE)
```
# Annex B - Effects of cleaning and grouping events
Does the grouping and cleanup of events that we did earlier have any noticeable effect?

```{r data-cleaning-1, cache=FALSE}
health.impact <- stormdata$FATALITIES + stormdata$INJURIES;
make_list(evgroup, health.impact)
```

- We did not see the `OTHER` group showing off, so the impact of the related events, even cumulated, shall be marginal.

If we list the impact relative to the original events:
```{r data-cleaning-2, cache=FALSE}
make_list(stormdata$EVTYPE, health.impact)
```
We can see that:

- TORNADO still dominates (probably because, by being the most frequent event, it is *de facto* better normalized).

- Heat is represented twice, as HEAT and as EXCESSIVE HEAT, so it was worth grouping them together. Similar comment for thunderstorm wind.

As a conclusion, the grouping and cleanup of event types improves the consistency of the results.
